### 2018-04-16 contents/reviews/trigeminal.tex
Updates from ShareLaTeX

The sensory afferents include the myelinated A and unmyelinated C-fibers that originate from the subcutaneous sensory receptors responsible for mechano- and temperature pain, fine touch and vibration, such as the Messner's, Pacinian and Ruffini corpuscles.

@@

V3 contains sensory and motor branches. It innervates skins of the jaw, areas above the ear, lips, gum, and teeth. The V3 division carries the mastication efferent to the four muscles recruited for mastications: the masseter, temporal, medial and lateral pterygoids. The tensor tympani also receives some CN V innervation, but it is primarily innervated by the facial nerve.  It is thought that CNV innervation is important in dampening the sound of chewing. V3 exits the skull through foramen ovale.

@@

The extension of the posterior aspect of the TRG forms the trigeminal nerve body, with both sensory and motor roots. The nerve body straddles the TRG and the trigeminal nucleus, by crossing the pontine cistern. The portion of the sensory nerve that enters the pons is termed the trigeminal root entry zone (REZ). The cistern trigeminal sensory nerve, similar to other peripheral nerves, is myelinated by Schwann cells. The myelin cell type transitions to oligodendrocyte myelinations in the CNS as it enters the pons. The CNS myelin can be several millimetres beyond the surface of the pons, before it gives away to the transition area, termed root entry zone, or Obersteiner-Redlich zone \cite{Peker2006}. Study in 100 CN V samples from 50 cadavers showed that cisternal CN V length ranges from 8-15 mm; the transition zone length is shorter on the medial aspect (0.1-2.5 mm) than the lateral aspect (0.17-6.75 mm). The transition zone can be up to 25\% of the nerve length \cite{Peker2006}.

@@

CN V passes into the pons (\ref{fig:trig-pathway}), through the transverse pontine fibers and enters the tegmentum of the pontine brainstem. It then curve towards the medial line before synapses onto the trigeminal nucleus in the basis of the pontine brainstem. The trigeminal nucleus is divided, from superior to inferior, into the mesencephalic, main sensory, and spinal nuclei. The spinal nucleus can be further divided into pars oralis, pars interpolaris and pars caudalis.

@@

The main sensory nucleus receives discriminative somatosensation from the oralfacial region, and along with the spinal trigeminal nucleus fibers, project to the contralateral ventral caudal nucleus (VC) of thalamus by decussating at various levels of the pons via the anterior trigeminothalamic tract. A separate branch responsible for oral cavity, however, projects to the ipsilateral VC via the dorsal trigeminothalamic tract.

@@

Classic TN is believed to be caused by neurovascular compression (NVC) of the CN V at its root entry zone (REZ) by the surrounding vasculature, where the nerve interfaces with the pons. An aberrant loop of artery and less commonly a vein is seen in 80-90\% of the cases of TN compressing the nerve \cite{Love2001,McLaughlin1999}. This can also be observed in conventional MR imaging \cite{Borges2010}. The superior cerebellar artery is responsible for most of the NVC (60-90\%), followed by the anterior inferior cerebellar artery (AICA) and basilar arteries \cite{Lutz2011}. There is a slight lateral prevalence, with right-sided TN occurring in  in 60\% of the cases, and in rare cases may be related to asymmetries of the foramen rotundum and foramen ovale \cite{Toda2009}.  Other types of compression may also cause TN; these can be in the forms of vestibular schwannomas, meningiomas, epidermoid cysts, and other cerebellopontine angle tumours, arteriovenous malformations and saccular aneurysms \cite{Haller2016,Love2001}. However NVC cannot be assumed to be directly causal of TN, since observation of vascular compression does not necessarily result in the clinical expression of pain \cite{Desouza2013,Hodaie2013}. In approximately 10-15\% of cases where TN occurs with no observable vascular compression \cite{Revuelta-Gutierrez2006,Maarbjerg2015}.

@@

Application of tractography permitted visual reconstructions of human cranial nerves in vivo \cite{Hodaie2010}. Cisternal CN V could be reliably delineated with Gaussian tensors with just 25 MR gradient directions. This permitted high fidelity visualization and targeted characterization of CN V diffusivity in TN, demonstrating significant reduction in fractional anisotropy (FA) post-surgery at the site of the GK target \cite{Hodaie2009a}. Study of CN V diffusivity in conjunction with deep white matter using TBSS, again revealed reduced FA on the affected CN V, with also bilateral increased in radial, axial, and mean diffusivity. Disruptions were also found in the cognitive-affective, attention, and motor regions of the cortex, consistent with cortical patterns in chronic pain \cite{Desouza2013}. Lowered FA was reported in multiple TN studies \cite{Herweh2007, Lutz2011}. Study in temporomandibular disorder (TMD) showed a bilateral reduction in CN V FA compared to controls, with the FA of the affected side significantly inverse-correlates with pain duration \cite{Moayedi2012}, thereby demonstrating that CN V FA is related to nociception. The TMD study also revealed similar cognitive and motor disruption, and internal capsule FA significantly inverse-correlated with pain intensity.

>>>>

Trigeminal sensory afferents include the myelinated A and unmyelinated C-fibers that originate from the subcutaneous sensory receptors responsible for mechano- and temperature pain, fine touch and vibration, such as the Messiner's, Pacinian and Ruffini corpuscles.

@@

V3 contains sensory and motor branches. It innervates skin of the jaw, areas above the ear, lips, gum, and teeth. The V3 division carries the mastication efferent to the four muscles recruited for mastications: the masseter, temporal, medial and lateral pterygoids. The tensor tympani also receives some CN V innervation, but it is primarily innervated by the facial nerve.  It is thought that CNV innervation is important in dampening the sound of chewing. V3 exits the skull through foramen ovale.

@@

The extension of the posterior aspect of the TRG forms the trigeminal nerve body, with both sensory and motor roots. The nerve body straddles the TRG and the trigeminal nucleus, by crossing the pontine cistern. The portion of the sensory nerve that enters the pons is termed the trigeminal root entry zone (REZ). The cistern trigeminal sensory nerve, similar to other peripheral nerves, is myelinated by Schwann cells. The myelin cell type transitions to oligodendrocyte myelinations in the CNS as it enters the pons. The CNS myelin can be several millimetres beyond the surface of the pons, before giving away to the transition area, termed root entry zone, or Obersteiner-Redlich zone \cite{Peker2006}. A study in 100 CN V samples from 50 cadavers showed that cisternal CN V length ranges from 8-15 mm; the transition zone length is shorter on the medial aspect (0.1-2.5 mm) than the lateral aspect (0.17-6.75 mm). The transition zone can be up to 25\% of the nerve length \cite{Peker2006}.

@@

CN V passes into the pons (\ref{fig:trig-pathway}), through the transverse pontine fibers and enters the tegmentum of the pontine brainstem. It then curve towards the medial aspect, and synapses onto the trigeminal nucleus in the basis of the pontine brainstem. The trigeminal brainstem sensory nucleus complex is divided, from superior to inferior, into the mesencephalic, main sensory, and spinal nuclei. The spinal nucleus can be further divided into pars oralis, pars interpolaris and pars caudalis.

@@

The main sensory nucleus receives discriminative input from the orofacial region, and along with the spinal trigeminal nucleus fibers, project to the contralateral ventral caudal nucleus (VC) of thalamus by decussating at various levels of the pons via the anterior trigeminothalamic tract. A separate branch responsible for oral cavity, however, projects to the ipsilateral VC via the dorsal trigeminothalamic tract.

@@

Classic TN is believed to be caused by neurovascular compression (NVC) of the CN V at its root entry zone (REZ) by the surrounding vasculature, where the nerve interfaces with the pons. An aberrant loop of artery and less commonly a vein is seen in 80-90\% of the cases of TN compressing the nerve \cite{Love2001,McLaughlin1999}. This can also be observed in conventional MR imaging \cite{Borges2010}. The superior cerebellar artery is responsible for most of the NVC (60-90\%), followed by the anterior inferior cerebellar artery (AICA) and basilar arteries \cite{Lutz2011}. There is a slight lateral prevalence, with right-sided TN occurring in  in 60\% of the cases, and in rare cases may be related to asymmetries of the foramen rotundum and foramen ovale \cite{Toda2009}.  Other types of compression may also cause TN; these can be in the forms of vestibular schwannomas, meningiomas, epidermoid cysts, and other cerebellopontine angle tumours, arteriovenous malformations and saccular aneurysms \cite{Haller2016,Love2001}. However NVC cannot be assumed to be directly causal of TN, since presence of vascular compression does not necessarily result in pain \cite{Desouza2013,Hodaie2013}. In approximately 10-15\% of cases where TN occurs with no observable vascular compression \cite{Revuelta-Gutierrez2006,Maarbjerg2015}.

@@

Application of tractography permitted visual reconstructions of human cranial nerves in vivo \cite{Hodaie2010}. Cisternal CN V could be reliably delineated with Gaussian tensors with just 25 MR gradient directions. This permitted high fidelity visualization and targeted characterization of CN V diffusivity in TN, demonstrating significant reduction in fractional anisotropy (FA) post-surgery at the site of the GK target \cite{Hodaie2009a}. Study of CN V diffusivity in conjunction with deep white matter again revealed reduced FA on the affected CN V, with also bilateral increased in radial, axial, and mean diffusivity. Disruptions were also found in the cognitive-affective, attention, and motor regions of the cortex, consistent with cortical patterns in chronic pain \cite{Desouza2013}. Lowered FA was reported in multiple TN studies \cite{Herweh2007, Lutz2011}. A Study in patients with temporomandibular disorder (TMD) showed a bilateral reduction in CN V FA compared to controls, with the FA of the affected side significantly inverse-correlates with pain duration \cite{Moayedi2012}. This TMD study also revealed similar cognitive and motor disruption, and internal capsule FA significantly inverse-correlated with pain intensity.
